Vagueness, Noise, and Communication

### 2014-10-02 paper/01_intro.tex
intro; sec com

>>>>

Many of our concepts and words are vague. A vague category knows clear
cases that fall under it, clear cases that do not, and also so-called
borderline cases in-between that neither clearly apply, nor clearly
not apply, but might themselves not be equal in terms of how well they
represent the category in question. Vagueness does not seem to
dramatically affect the success of everyday communication, but it is
troublesome for some of our best theories of language. Especially the
logico-semantic tradition of Frege, Russell and the young Wittgenstein
is challenged by vagueness and the paradoxes that it gives rise to. As
can be expected, many proponents of this tradition have risen to the
challenge \citep{Williamson1994:Vagueness}.

There are other intriguing aspects about vagueness. Linguists, for
example, wonder about the reasons why some expressions are less vague
than others \citep[e.g.][]{Kennedy2007:Vagueness-and-G}. But the
puzzle that we are concerned with here is how vagueness could arise
and be maintained in the first place. This may not seem a particularly
deep issue at first glance, but on closer inspection it is a serious
worry to functionalist accounts that maintain that our concepts and
language use evolved in order to be efficient. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
exist despite its obvious and demonstrable functional deficiency under
evolutionary pressure to be optimally expressive.

This problem, call it Lipman's problem, has a conceptual and a
technical side to it. Conceptually, it asks for reasons and general
mechanisms by which we could plausibly conceive of vagueness as
persisting the pressure of evolutionary optimization towards
precision. Such reasons and general mechanisms are, arguably, not hard
to imagine. We will address some presently. But each putative
explanation, no matter how plausible intuitively, must also be checked
for internal consistency and its potential to account for the fact
that we see vagueness as a part of a by-and-large efficient system of
categorization and communication. In other words, not all first-shot
rebuttals of Lipman's problem will do. Denying that there is any
functional pressure on efficient communication, for instance, is
dubious because it leaves the relative efficiency of our communication
with vague words entirely unexplained. In sum, the technical answer to
Lipman's problem is to give, ideally, a model of evolution of
categorization or language use that shows how competing forces that
realize both by-and-large efficient categories and vagueness.

c

\bigskip
